
---
title: "MAGeCKFlute GSEA and R1–R2 Correlation Analysis"
author: "CLN5 Batten Disease Genetic Modifier Screen"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This workflow performs gene set enrichment analysis (GSEA) on MAGeCK-MLE
gene-level beta scores and assesses concordance between R1 and R2 results.

Four GSEA analyses are performed:
- R1 Upper Lysotracker (UL) versus D7
- R1 Lower Lysotracker (LL) versus D7
- R2 UL versus D7
- R2 LL versus D7

Non-targeting controls are excluded from pathway analysis. Pearson
correlations compare R1 and R2 beta scores for UL and LL separately.

## 1. Setup


```{r setup, message=FALSE, warning=FALSE}
library(MAGeCKFlute)
library(dplyr)
library(ggplot2)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(knitr)
library(readxl)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Load R1 and R2 MAGeCK2-MLE results
R1_data <- read_excel(
  "R1_MLE_Diff20_median.gene_summary.xlsx"
)

R2_data <- read_excel(
  "R2_MAGECK2_Diff20_median.gene_summary.txt.xlsx"
)

# Create output directories
dir.create("results", showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# Check expected columns
required_cols <- c("Gene", "Lower|beta", "Upper|beta")

stopifnot(all(required_cols %in% colnames(R1_data)))
stopifnot(all(required_cols %in% colnames(R2_data)))

# Exclude non-targeting controls
R1_data <- R1_data %>%
  filter(Gene != "non-targeting")

R2_data <- R2_data %>%
  filter(Gene != "non-targeting")

# Summarise input gene counts
data.frame(
  Dataset = c("R1", "R2"),
  Genes_after_NTC_exclusion = c(
    nrow(R1_data),
    nrow(R2_data)
  )
) %>%
  kable(
    caption = "Input gene counts after excluding non-targeting controls"
  )

```

## 2. Map gene symbols to Entrez IDs

The GSEA workflow uses Entrez identifiers for compatibility with the
gene sets used by MAGeCKFlute. Symbols that map to multiple Entrez IDs
are excluded to avoid ambiguous mappings. If multiple symbols map to
the same Entrez ID, the symbol with the largest absolute beta score is
retained for that comparison.

```{r gene-mapping}
prepare_rank <- function(data, beta_column) {

  genes <- data %>%
    transmute(
      SYMBOL = as.character(Gene),
      beta = as.numeric(.data[[beta_column]])
    ) %>%
    filter(
      !is.na(SYMBOL),
      SYMBOL != "",
      is.finite(beta)
    ) %>%
    distinct(SYMBOL, .keep_all = TRUE)

  input_symbols <- unique(genes$SYMBOL)

  # Map recognised gene symbols directly.
  direct_mapping <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = input_symbols,
    keytype = "SYMBOL",
    columns = c("SYMBOL", "ENTREZID")
  ) %>%
    filter(!is.na(ENTREZID)) %>%
    transmute(
      input_SYMBOL = SYMBOL,
      SYMBOL,
      ENTREZID = as.character(ENTREZID)
    ) %>%
    distinct()

  # Identify symbols not recognised by the direct SYMBOL mapping.
  unmapped_symbols <- setdiff(
    input_symbols,
    direct_mapping$input_SYMBOL
  )

  # Try mapping unrecognised symbols as aliases.
  alias_mapping <- if (length(unmapped_symbols) > 0) {
    AnnotationDbi::select(
      org.Hs.eg.db,
      keys = unmapped_symbols,
      keytype = "ALIAS",
      columns = c("ALIAS", "SYMBOL", "ENTREZID")
    ) %>%
      filter(!is.na(ENTREZID)) %>%
      transmute(
        input_SYMBOL = ALIAS,
        SYMBOL,
        ENTREZID = as.character(ENTREZID)
      ) %>%
      distinct()
  } else {
    data.frame(
      input_SYMBOL = character(),
      SYMBOL = character(),
      ENTREZID = character()
    )
  }

  mapping <- bind_rows(direct_mapping, alias_mapping) %>%
    distinct(input_SYMBOL, SYMBOL, ENTREZID)

  # Exclude input symbols that map to multiple Entrez IDs.
  ambiguous_symbols <- mapping %>%
    group_by(input_SYMBOL) %>%
    summarise(n_ids = n_distinct(ENTREZID), .groups = "drop") %>%
    filter(n_ids > 1) %>%
    pull(input_SYMBOL)

  mapping_unique <- mapping %>%
    filter(!input_SYMBOL %in% ambiguous_symbols) %>%
    distinct(input_SYMBOL, ENTREZID)

  # Exclude every Entrez ID associated with multiple input symbols.
  duplicated_entrez <- mapping_unique %>%
    count(ENTREZID, name = "n_symbols") %>%
    filter(n_symbols > 1) %>%
    pull(ENTREZID)

  mapping_clean <- mapping_unique %>%
    filter(!ENTREZID %in% duplicated_entrez)

  mapped <- genes %>%
    inner_join(
      mapping_clean,
      by = c("SYMBOL" = "input_SYMBOL")
    ) %>%
    mutate(ENTREZID = as.character(ENTREZID))

  rank_vector <- mapped$beta
  names(rank_vector) <- mapped$ENTREZID
  rank_vector <- sort(rank_vector, decreasing = TRUE)

  list(
    rank = rank_vector,
    mapping = mapping,
    ambiguous_symbols = ambiguous_symbols,
    duplicated_entrez = duplicated_entrez,
    mapped_genes = mapped,
    input_genes = nrow(genes),
    mapped_unique_genes = length(rank_vector)
  )
}

R1_UL <- prepare_rank(R1_data, "Upper|beta")
R1_LL <- prepare_rank(R1_data, "Lower|beta")
R2_UL <- prepare_rank(R2_data, "Upper|beta")
R2_LL <- prepare_rank(R2_data, "Lower|beta")

mapping_QC <- data.frame(
  Analysis = c("R1 UL", "R1 LL", "R2 UL", "R2 LL"),
  Input_genes = c(
    R1_UL$input_genes, R1_LL$input_genes,
    R2_UL$input_genes, R2_LL$input_genes
  ),
  Unique_Entrez_genes = c(
    R1_UL$mapped_unique_genes, R1_LL$mapped_unique_genes,
    R2_UL$mapped_unique_genes, R2_LL$mapped_unique_genes
  ),
  Ambiguous_symbols = c(
    length(R1_UL$ambiguous_symbols),
    length(R1_LL$ambiguous_symbols),
    length(R2_UL$ambiguous_symbols),
    length(R2_LL$ambiguous_symbols)
  )
)

write.csv(
  mapping_QC,
  "results/gene_mapping_QC.csv",
  row.names = FALSE
)

kable(mapping_QC, caption = "Gene identifier mapping QC")
```

## 3. Run the four GSEA analyses

Genes are ranked by signed beta score in descending order. Positive
beta scores therefore occur at the top of the list and negative beta
scores at the bottom.

GSEA tests Gene Ontology Biological Process (GOBP) gene sets using
MAGeCKFlute's `enrich.GSE()` function. A p-value cutoff of 1 retains
the full results table for downstream FDR filtering.

```{r run-gsea}
run_gsea <- function(rank_vector) {
  MAGeCKFlute::enrich.GSE(
    rank_vector,
    keytype = "Entrez",
    type = "GOBP",
    organism = "hsa",
    pvalueCutoff = 1
  )
}

R1_UL_GSEA <- run_gsea(R1_UL$rank)
R1_LL_GSEA <- run_gsea(R1_LL$rank)
R2_UL_GSEA <- run_gsea(R2_UL$rank)
R2_LL_GSEA <- run_gsea(R2_LL$rank)

gsea_results <- list(
  R1_UL = as.data.frame(R1_UL_GSEA),
  R1_LL = as.data.frame(R1_LL_GSEA),
  R2_UL = as.data.frame(R2_UL_GSEA),
  R2_LL = as.data.frame(R2_LL_GSEA)
)

# Save complete pathway tables, including nonsignificant pathways.
for (nm in names(gsea_results)) {
  write.csv(
    gsea_results[[nm]],
    file.path("results", paste0(nm, "_GSEA_all_pathways.csv")),
    row.names = FALSE
  )
}

gsea_QC <- data.frame(
  Analysis = names(gsea_results),
  Genes_in_rank = c(
    length(R1_UL$rank), length(R1_LL$rank),
    length(R2_UL$rank), length(R2_LL$rank)
  ),
  Pathways_returned = vapply(gsea_results, nrow, integer(1)),
  NA_pvalues = vapply(
    gsea_results,
    function(x) sum(is.na(x$pvalue)),
    integer(1)
  ),
  FDR_below_0.05 = vapply(
    gsea_results,
    function(x) sum(x$p.adjust < 0.05, na.rm = TRUE),
    integer(1)
  )
)

write.csv(
  gsea_QC,
  "results/GSEA_QC_summary.csv",
  row.names = FALSE
)

kable(gsea_QC, caption = "GSEA analysis QC summary")
```

## 4. Extract candidate pathways

Candidate enriched pathways are defined using the thesis criteria:
FDR-adjusted p-value below 0.05 and normalised enrichment score (NES)
greater than 1. Depleted pathways with NES below -1 are also saved
separately for exploratory interpretation.

```{r significant-pathways}
candidate_pathways <- lapply(names(gsea_results), function(nm) {
  gsea_results[[nm]] %>%
    filter(!is.na(p.adjust), !is.na(NES),
           p.adjust < 0.05, NES > 1) %>%
    mutate(Analysis = nm)
}) %>%
  bind_rows()

depleted_pathways <- lapply(names(gsea_results), function(nm) {
  gsea_results[[nm]] %>%
    filter(!is.na(p.adjust), !is.na(NES),
           p.adjust < 0.05, NES < -1) %>%
    mutate(Analysis = nm)
}) %>%
  bind_rows()

write.csv(
  candidate_pathways,
  "results/significant_enriched_pathways.csv",
  row.names = FALSE
)

write.csv(
  depleted_pathways,
  "results/significant_depleted_pathways.csv",
  row.names = FALSE
)


kable(
  candidate_pathways %>%
    dplyr::select(
      Analysis, ID, Description, NES, pvalue, p.adjust, Count
    ),
  caption = "Significant enriched pathways (FDR < 0.05 and NES > 1)"
)
```

## 5. Pearson correlation of R1 and R2 beta scores

R1 and R2 results are paired by gene symbol. Non-targeting controls
and genes without a finite beta score in either dataset are excluded.
Pearson correlation assesses the linear concordance of gene-level
effects between sequencing read orientations; it does not establish
causation.

```{r pearson-correlations}
compare_beta <- function(column, population_label) {

  
paired <- R1_data %>%
  dplyr::select(
    Gene,
    R1_beta = all_of(column)
  ) %>%
  dplyr::inner_join(
    R2_data %>%
      dplyr::select(
        Gene,
        R2_beta = all_of(column)
      ),
    by = "Gene"
  ) %>%
  dplyr::mutate(
    R1_beta = as.numeric(R1_beta),
    R2_beta = as.numeric(R2_beta)
  ) %>%
  dplyr::filter(
    is.finite(R1_beta),
    is.finite(R2_beta)
  )

  test <- cor.test(
    paired$R1_beta,
    paired$R2_beta,
    method = "pearson"
  )

  plot <- ggplot(paired, aes(x = R1_beta, y = R2_beta)) +
    geom_point(alpha = 0.55, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE, colour = "black") +
    theme_classic(base_size = 12) +
    labs(
      title = paste("R1 versus R2 beta scores:", population_label),
      subtitle = paste0(
        "Pearson r = ", round(unname(test$estimate), 3),
        "; p = ", signif(test$p.value, 3),
        "; n = ", nrow(paired), " genes"
      ),
      x = "R1 gene-level beta score",
      y = "R2 gene-level beta score"
    )

  ggsave(
    filename = file.path(
      "figures",
      paste0("Pearson_R1_R2_", population_label, ".png")
    ),
    plot = plot,
    width = 7,
    height = 6,
    dpi = 300
  )

  list(
    plot = plot,
    summary = data.frame(
      Population = population_label,
      Genes_compared = nrow(paired),
      Pearson_r = unname(test$estimate),
      P_value = test$p.value,
      CI_lower = test$conf.int[1],
      CI_upper = test$conf.int[2]
    ),
    paired_data = paired
  )
}

pearson_UL <- compare_beta("Upper|beta", "UL")
pearson_LL <- compare_beta("Lower|beta", "LL")

pearson_summary <- bind_rows(
  pearson_UL$summary,
  pearson_LL$summary
)

write.csv(
  pearson_summary,
  "results/Pearson_R1_R2_correlation_summary.csv",
  row.names = FALSE
)

write.csv(
  pearson_UL$paired_data,
  "results/Pearson_R1_R2_paired_beta_UL.csv",
  row.names = FALSE
)

write.csv(
  pearson_LL$paired_data,
  "results/Pearson_R1_R2_paired_beta_LL.csv",
  row.names = FALSE
)

pearson_UL$plot
pearson_LL$plot

kable(
  pearson_summary,
  digits = 4,
  caption = "Pearson correlations between R1 and R2 gene-level beta scores"
)
```

## 6. Session information

Record software versions to support reproducibility.

```{r session-info}
sessionInfo()
```




